---
title: "Statistical Models for Networks "
subtitle: "Inferential approaches to network structure"
author: "Dr Clemens Jarnach"
affiliation: "Oxford Martin School, University of Oxford"
date: "03-18-2026"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: cosmo
    self-contained: true
execute:
  warning: false
  message: false
---

Descriptive measures tell us what networks look like; statistical models let us test hypotheses about *why* they look that way. This chapter introduces inferential frameworks for networks, focusing on **Exponential Random Graph Models (ERGMs)** — the workhorse of statistical network analysis in the social sciences.

---

## Why Standard Regression Fails for Networks

Before diving into ERGMs, it is worth understanding why we cannot simply regress a tie indicator (edge = 1, no edge = 0) on node or dyad attributes using OLS or logistic regression.

The core problem is **non-independence**. In standard regression, we assume that each observation is independent of every other. In a network, that assumption is violated almost by definition:

- Whether Alice is friends with Bob likely depends on whether Alice is friends with Carol, and whether Carol is friends with Bob (transitivity / triadic closure).
- A node with many friends is more likely to acquire *even more* friends (degree-based effects / preferential attachment).
- Ties between pairs of nodes are not drawn independently — the whole point of network analysis is that structure is relational and recursive.

If you ignore this and run a naive logistic regression on dyad-level tie indicators, your standard errors are likely wrong, and you may overstate statistical significance. The network is not a bag of independent coin flips — it is a *system*.

**Exponential Random Graph Models** (ERGMs, also called p* models) address this directly. An ERGM defines a probability distribution over the entire space of possible networks of a given size. It asks: given some set of network statistics (edge count, triangles, degree distribution, nodal covariates), what is the probability of observing *this particular network*?

Formally, an ERGM specifies:

$$P(Y = y) = \frac{\exp\left(\boldsymbol{\theta}^\top \mathbf{g}(y)\right)}{\kappa(\boldsymbol{\theta})}$$

where:

- $Y$ is the random network (each $Y_{ij}$ is a potential tie),
- $y$ is the observed network,
- $\mathbf{g}(y)$ is a vector of network statistics (e.g., number of edges, triangles, attribute-based counts),
- $\boldsymbol{\theta}$ is the corresponding vector of parameters to be estimated,
- $\kappa(\boldsymbol{\theta})$ is a normalising constant summing over *all possible networks* of the same size — which is what makes these models computationally demanding (and is why estimation uses MCMC).

Positive $\theta$ for a statistic means that networks with more of that feature are more likely than chance; negative means fewer. This gives us an intuitive, hypothesis-driven framework for asking: do triangles appear more often than we would expect at random? Do wealthy nodes preferentially connect to one another?

---

## Setup

```{r}
#| message: false
#| warning: false

# Load only the statnet components needed for ERGMs
# (avoids dependency on tergm and other optional suite packages)
library(network)   # network objects
library(ergm)      # ERGM estimation and GoF
library(sna)       # network descriptives (gden, etc.)
library(ggplot2)   # for any custom plotting
```

---

## The Florentine Families Dataset

We will work with the **Florentine Families** dataset — a classic in social network analysis, first described by Padgett & Ansell (1993). It captures **marriage ties** among 16 prominent Florentine families in 15th-century Renaissance Italy. The Medici family's rise to political dominance is a central story in the dataset: their strategic position in the marriage network gave them outsized influence.

The dataset is bundled with `ergm` (part of `statnet`) as `flomarriage`.

```{r}
# Load the dataset
# Load the florentine families data
data(florentine)
data(flomarriage)

# Inspect the network object
flomarriage
```

```{r}
# Node-level attributes
flomarriage %v% "wealth"     # Annual income in thousands of lira (1427)
flomarriage %v% "priorates"  # Number of seats on city council 1282-1344
flomarriage %v% "vertex.names"
```

```{r}
# Basic network descriptives
network.size(flomarriage)        # Number of nodes
network.edgecount(flomarriage)   # Number of edges
gden(flomarriage)                # Density
```

The network has 16 nodes (families) and 20 marriage ties. With a density of roughly 0.17, ties are sparse — typical of elite social networks where strategic marriages were carefully chosen, not randomly distributed.

```{r}
# Visualise the network with wealth encoded in node size
par(mar = c(1, 1, 2, 1))

plot(
  flomarriage,
  vertex.cex    = flomarriage %v% "wealth" / 30,
  vertex.col    = "steelblue",
  edge.col      = "grey50",
  label         = flomarriage %v% "vertex.names",
  label.cex     = 0.7,
  label.pos     = 5,
  main          = "Florentine Marriage Network\n(node size = wealth)"
)
```

The Medici sit at the structural heart of this network. Despite not being the wealthiest family, they bridged otherwise disconnected clusters — a textbook example of brokerage and betweenness centrality translating into political power.

---

## Model 1: The Bernoulli (Null) Model

We start with the simplest possible ERGM: a **Bernoulli graph** or *Erdős–Rényi* model. The only term is `edges`, which acts like an intercept — it captures the baseline probability of a tie forming, and nothing else.

```{r}
model_null <- ergm(flomarriage ~ edges)
summary(model_null)
```

The coefficient for `edges` is negative, which makes sense: most possible ties are *absent* (the network is sparse). This log-odds can be converted to a probability:

```{r}
# Tie probability under null model
plogis(coef(model_null)["edges"])
```

This matches the network density exactly — as it should. The null model is just a fancy way of saying "each tie forms independently with probability equal to the overall density." It is our baseline: every more complex model will be judged against it.

---

## Model 2: Structural Effects — Transitivity and GWESP

Real social networks are not Bernoulli. One of the most robust findings in network science is **triadic closure**: if A is tied to B, and B is tied to C, then A and C are more likely to be tied too. Friends of friends become friends. In the Florentine context: if the Medici married into the Tornabuoni *and* the Tornabuoni are linked to the Strozzi, the Medici-Strozzi tie becomes more likely.

The naive way to capture this is the `triangle` term — counting the number of closed triangles in the network. However, `triangle` is notorious for causing **model degeneracy**: the ERGM collapses onto a near-complete or near-empty graph, producing nonsensical estimates. This is a known pathology of the raw triangle term.

The solution is the **Geometrically Weighted Edgewise Shared Partners** (GWESP) statistic, introduced by Hunter & Handcock (2006). GWESP counts shared partners but applies a geometric decay: the *first* shared partner between a dyad contributes fully, the *second* contributes less, and so on. This prevents the model from being dominated by a few highly-triangulated nodes and greatly improves stability.

```{r}
model_structural <- ergm(
  flomarriage ~ edges + triangle + gwesp(0.5, fixed = TRUE),
  control = control.ergm(seed = 42)
)
summary(model_structural)
```

> **Note on `triangle`:** Including `triangle` alongside `gwesp` is illustrative here — in practice you would typically use one or the other (preferring `gwesp`). If you encounter degeneracy warnings, drop `triangle` and rely solely on `gwesp`.

```{r}
# Fit GWESP-only structural model (more stable)
model_gwesp <- ergm(
  flomarriage ~ edges + gwesp(0.5, fixed = TRUE),
  control = control.ergm(seed = 42)
)
summary(model_gwesp)
```

A positive and significant `gwesp` coefficient would tell us that triadic closure is a real force in this network — dyads that share at least one partner are more likely to be tied. However, the results are not statistical significant, so we may need to dig deeper. 

---

## Model 3: Homophily — Does Wealth Drive Tie Formation?

Network structure is shaped not only by internal dynamics (transitivity, degree) but also by **node attributes**. A key question for the Florentine data: did wealthier families preferentially form marriage ties?

We test two hypotheses:

1. **Wealth activity effect** (`nodecov`): Do wealthier families have *more ties in general*? A positive `nodecov` coefficient means that nodes with higher wealth values are more likely to be involved in any given dyad being a tie.

2. **Wealth homophily** (`absdiff`): Do families of *similar wealth* preferentially marry one another? The `absdiff` term computes the absolute difference in wealth between each pair of nodes; a *negative* coefficient would indicate homophily (smaller differences → more ties).

```{r}
model_wealth <- ergm(
  flomarriage ~ edges + gwesp(0.5, fixed = TRUE) + nodecov("wealth") + absdiff("wealth"),
  control = control.ergm(seed = 42)
)
summary(model_wealth)
```

Interpreting the results:

- A positive `nodecov("wealth")` suggests wealthier families are more active in forming ties overall.
- A negative `absdiff("wealth")` would indicate wealth homophily — families with similar wealth are drawn together.
- We evaluate statistical significance through the z-scores (estimate / std. error) and p-values in the summary output.

---

## Model Comparison: AIC and BIC

With multiple models in hand, we need a principled basis for comparison. **AIC** (Akaike Information Criterion) and **BIC** (Bayesian Information Criterion) penalise model complexity to guard against overfitting: lower is better.

```{r}
# Extract AIC and BIC for each model
model_list <- list(
  Null       = model_null,
  GWESP      = model_gwesp,
  Wealth     = model_wealth
)

comparison_table <- data.frame(
  Model  = names(model_list),
  AIC    = sapply(model_list, AIC),
  BIC    = sapply(model_list, BIC),
  LogLik = sapply(model_list, logLik)
)

comparison_table
```

```{r}
# Tidy comparison with knitr (renders nicely in HTML)
knitr::kable(
  comparison_table,
  digits  = 2,
  caption = "Model comparison: Florentine marriage network ERGMs"
)
```

A model with substantially lower AIC/BIC than the null model would suggest that the added structural or covariate terms meaningfully improve fit beyond what chance predicts.

In our example, the wealth covariates provide the only substantive improvement in fit (log-likelihood gain of ~3.3), and AIC marginally favours including them; but the effect is small. Moreover, GWESP does essentially nothing here. Triadic closure may simply not be a strong structural force in this small, sparse network, or, the likely reason here, the network is too small to estimate it reliably. BIC's preference for the Null is largely a function of the tiny sample (n = 16, 20 edges) — BIC is very conservative in small networks.
The honest conclusion: with 16 nodes, you do not have the statistical power to decisively distinguish these models on AIC/BIC alone. 

---

## MCMC Diagnostics

ERGMs are estimated using **Markov Chain Monte Carlo (MCMC)**. The MCMC sampler proposes changes to a simulated network and accepts or rejects them to approximate the normalising constant $\kappa(\boldsymbol{\theta})$ that cannot be computed analytically.

Before trusting your estimates, you *must* check that the MCMC chains have converged. A non-converged chain means the parameter estimates are unreliable, regardless of how significant the z-scores look.

`mcmc.diagnostics()` produces trace plots and autocorrelation diagnostics for each parameter. What you want to see:

- **Trace plots** that look like "fuzzy caterpillars" — bouncing randomly around a stable mean with no obvious trends or drift.
- **Low autocorrelation** — the chain is mixing well and not getting stuck.
- **Sample statistics close to observed** — the mean of the simulated statistics should match the observed statistics.

```{r}
#| fig-width: 8
#| fig-height: 6

mcmc.diagnostics(model_wealth)
```

If you see systematic drift, poor mixing, or large discrepancies between simulated and observed statistics, you may need to adjust MCMC settings (longer chains, thinner, more burnin) via `control.ergm()`.

---

## Goodness of Fit

Even if MCMC has converged, we need to check whether the model actually *fits* the observed network. **Goodness of Fit (GoF)** works as follows:

1. Simulate a large number of networks from the fitted ERGM.
2. Compute key network statistics for each simulated network.
3. Compare those distributions to the corresponding statistics of the *observed* network.

If the observed statistics fall within the bulk of the simulated distribution, the model fits. If they fall in the tails, the model is missing something important.

The standard GoF diagnostic in `ergm` checks four statistics:

- **Degree distribution**: does the model reproduce the distribution of node degrees?
- **Edgewise shared partners (ESP)**: does the model reproduce local clustering structure?
- **Geodesic distance distribution**: does the model reproduce path length structure?
- **Model statistics**: do simulated networks reproduce the terms in the model?

```{r}
#| fig-width: 9
#| fig-height: 7

gof_wealth <- gof(model_wealth)
gof_wealth
```

```{r}
#| fig-width: 9
#| fig-height: 7

par(mfrow = c(2, 2))
plot(gof_wealth)
```

In each panel, the **black line** is the observed statistic; the **boxplots** summarise the distribution across simulated networks; the **grey band** is the 95% confidence envelope. Observed values that fall comfortably within the envelope indicate a well-fitting model.

Compare GoF for the null model:

```{r}
#| fig-width: 9
#| fig-height: 7

gof_null <- gof(model_null)
par(mfrow = c(2, 2))
plot(gof_null)
```

The null model typically fails on degree distribution and ESP — it cannot reproduce the clustering we see in the observed network. The richer models should show a marked improvement.

---

## Interpreting ERGM Coefficients

ERGM parameters are on the **log-odds** scale for individual tie changes, *holding the rest of the network fixed*. This "change statistic" interpretation is subtle but important:

- The `edges` coefficient is the log-odds of a tie forming between two nodes when no other network statistics would change — i.e., the baseline propensity.
- The `gwesp` coefficient is the *additional* log-odds of a tie forming for each additional (geometrically weighted) shared partner the dyad has.
- The `nodecov("wealth")` coefficient is the additional log-odds contribution per unit increase in the *sum* of the two nodes' wealth values.
- The `absdiff("wealth")` coefficient is the additional log-odds contribution per unit increase in the *absolute difference* in wealth.

```{r}
# Convert log-odds to odds ratios for easier interpretation
exp(coef(model_wealth))
```

An odds ratio > 1 increases the probability of a tie; < 1 decreases it. For example, if `exp(coef(model_wealth)["gwesp.fixed.0.5"])` is 2.5, then each additional (weighted) shared partner multiplies the odds of a tie by 2.5.

---

## Summary and Key Takeaways

ERGMs provide a principled, statistically rigorous framework for asking whether network structure arises from:

- **Random chance** (the null/Bernoulli model),
- **Endogenous structural forces** like transitivity and degree-based effects (GWESP),
- **Exogenous nodal attributes** like wealth (nodecov, absdiff, nodematch), or
- **Some combination** of all of the above.

For the Florentine marriage network, the evidence points to a network shaped by both structural closure and wealth-based sorting — elite families in Renaissance Florence did not marry randomly, and the resulting network topology reflected deliberate social strategy.

The workflow introduced here — **Specification → Estimation → MCMC Diagnostics → Goodness of Fit** — is the standard pipeline for any applied ERGM analysis. Never skip diagnostics: a model that has not converged or that fails GoF checks should not be interpreted, regardless of how tidy the summary table looks.

---

## Further Reading

- Lusher, Dean, Johan Koskinen, and Garry Robins. [-@lusherExponentialRandomGraph2013]. Exponential Random Graph Models for Social Networks : Theories, Methods, and Applications. Cambridge: Cambridge University Press.
- Rawlings, Craig M., Jeffrey A. Smith, James W. Moody, and Daniel McFarland. [-@rawlingsNetworkAnalysisIntegrating2023] Network Analysis : Integrating Social Network Theory, Method, and Application with R. Cambridge: Cambridge University Press. https://inarwhal.github.io/NetworkAnalysisR-book/.
- Padgett, J. F., & Ansell, C. K. (1993). Robust action and the rise of the Medici. *American Journal of Sociology*, 98(6), 1259–1319.
- The `statnet` project: [statnet.org](https://statnet.org) — documentation, tutorials, and workshops.

