---
title: "Network Layout Algorithm Comparison"
subtitle: "How visualisation choices shape what we see in a network"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    fig-width: 10
    fig-height: 7
execute:
  warning: false
  message: false
---

## Overview

A network layout algorithm takes a graph (nodes and edges) and assigns each node an `(x, y)` position on a 2-D canvas. The *same* network can look radically different depending on which algorithm you choose — and choosing the right one matters because it affects which structures are visually salient.

This notebook:

1.  Generates a synthetic **small-world network** to use as a shared test case.
2.  Runs **six common layout algorithms** side-by-side so you can compare their outputs directly.
3.  Dives deeper into the **Fruchterman-Reingold** algorithm and shows how its key parameters change the resulting picture.

------------------------------------------------------------------------

## Setup

```{r}
#| label: setup
library(igraph)
```

------------------------------------------------------------------------

## 1 · Generate a Small-World Network

We use the **Watts-Strogatz** model (`sample_smallworld`), a classic benchmark that produces networks with:

-   **High clustering** — neighbours of a node tend to know each other.
-   **Short average path lengths** — most pairs of nodes are only a few steps apart.

These properties are found in many real-world networks (social, biological, technological), making it a good test case for layout algorithms.

```{r}
#| label: generate-network

set.seed(42)   # fix randomness for reproducibility

# 200 nodes arranged on a 1-D ring; each node connected to its 3 nearest
# neighbours on each side; each edge independently rewired with prob 0.03
g <- sample_smallworld(dim = 1, size = 200, nei = 3, p = 0.03)
```

### Inspect the network

```{r}
#| label: network-summary

cat("Network summary:\n")
print(g)
cat("\nNodes:                  ", vcount(g), "\n")
cat("Edges:                  ", ecount(g), "\n")
cat("Average path length:    ", round(mean_distance(g), 3), "\n")
cat("Clustering coefficient: ", round(transitivity(g, type = "average"), 3), "\n")
```

> **Short average path length + high clustering** = hallmarks of a small-world network.

------------------------------------------------------------------------

## 2 · Shared Visual Style

To make a fair comparison between layouts we fix every visual parameter — colour, size, line width — and only vary the `layout` argument.

```{r}
#| label: visual-style

node_col   <- "#4682B4"   # steel blue
edge_col   <- "#AAAAAA"   # light grey
node_size  <- 5
edge_width <- 0.6
```

------------------------------------------------------------------------

## 3 · Compute All Layouts

We compute every layout **once** before plotting. This keeps run-times predictable and ensures each algorithm starts from the same underlying graph object.

```{r}
#| label: compute-layouts

layout_fr     <- layout_with_fr(g)     # Fruchterman-Reingold (force-directed)
layout_kk     <- layout_with_kk(g)     # Kamada-Kawai (spring-electrical)
layout_drl    <- layout_with_drl(g)    # DrL — designed for large graphs
layout_lgl    <- layout_with_lgl(g)    # Large Graph Layout
layout_circle <- layout_in_circle(g)   # Circular — purely deterministic
layout_random <- layout_randomly(g)    # Random — no optimisation (baseline)
```

------------------------------------------------------------------------

## 4 · Side-by-Side Layout Comparison

The six panels below show identical data — the same 200-node small-world graph — rendered with six different algorithms. Pay attention to which structures each layout makes easy (or hard) to see.

```{r}
#| label: layout-comparison
#| fig-cap: "Six layout algorithms applied to the same small-world network."
#| fig-height: 8

par(mfrow = c(2, 3), mar = c(1, 1, 2.5, 1))

plot(g,
     layout       = layout_fr,
     main         = "Fruchterman-Reingold",
     vertex.size  = node_size, vertex.color = node_col,
     vertex.label = NA, edge.color = edge_col, edge.width = edge_width)

plot(g,
     layout       = layout_kk,
     main         = "Kamada-Kawai",
     vertex.size  = node_size, vertex.color = node_col,
     vertex.label = NA, edge.color = edge_col, edge.width = edge_width)

plot(g,
     layout       = layout_drl,
     main         = "DrL",
     vertex.size  = node_size, vertex.color = node_col,
     vertex.label = NA, edge.color = edge_col, edge.width = edge_width)

plot(g,
     layout       = layout_lgl,
     main         = "Large Graph Layout (LGL)",
     vertex.size  = node_size, vertex.color = node_col,
     vertex.label = NA, edge.color = edge_col, edge.width = edge_width)

plot(g,
     layout       = layout_circle,
     main         = "Circle",
     vertex.size  = node_size, vertex.color = node_col,
     vertex.label = NA, edge.color = edge_col, edge.width = edge_width)

plot(g,
     layout       = layout_random,
     main         = "Random (baseline)",
     vertex.size  = node_size, vertex.color = node_col,
     vertex.label = NA, edge.color = edge_col, edge.width = edge_width)

par(mfrow = c(1, 1), mar = c(5, 4, 4, 2) + 0.1)
```

### Algorithm cheat-sheet

| Algorithm | Core idea | Best used when… |
|----|----|----|
| **Fruchterman-Reingold** | Nodes repel each other; edges act as springs pulling connected nodes together | General purpose; medium-sized graphs |
| **Kamada-Kawai** | Minimises the difference between graph-theoretic distance and geometric distance | Smaller graphs where geodesic distance should drive positions |
| **DrL** | Multilevel force-directed; groups nodes before placing them | Large graphs (thousands of nodes) |
| **LGL** | Builds outward from a root node using minimum spanning tree | Large, sparse graphs with clear hierarchy |
| **Circle** | Equally spaced nodes on a ring | Showing all nodes at equal visual weight; small graphs |
| **Random** | Positions drawn uniformly at random | Baseline comparison; never use in final visualisations |

------------------------------------------------------------------------

## 5 · Fruchterman-Reingold Parameter Tuning

Fruchterman-Reingold (FR) is one of the most widely used force-directed algorithms. Understanding its parameters lets you control the trade-off between **global structure** (which clusters are far from which?) and **local refinement** (are tightly connected nodes packed together?).

### Key parameters

| Parameter | Default | Effect |
|----|----|----|
| `niter` | 500 | Number of iterations the algorithm runs. More iterations = more time for the layout to settle. |
| `start.temp` | `sqrt(node count)` | Starting "temperature" — the maximum distance a node can move in a single iteration. |

**Temperature intuition:**

-   **High temperature** → nodes can jump large distances per step → the algorithm explores the full space broadly, revealing global structure.
-   **Low temperature** → nodes move in tiny steps → the algorithm barely moves from starting positions, only doing local fine-tuning.

### Compute six FR variants

```{r}
#| label: fr-variants

# 1. Default — reference point
fr_default <- layout_with_fr(g)

# 2. Very few iterations — layout hasn't had time to converge
fr_few_iter <- layout_with_fr(g, niter = 30)

# 3. Many iterations — fully converged, most refined
fr_many_iter <- layout_with_fr(g, niter = 5000)

# 4. High starting temperature — large displacements allowed;
#    long-range/global structure is prioritised
fr_high_temp <- layout_with_fr(g, start.temp = vcount(g) * 2)

# 5. Low starting temperature — nodes barely move from random start;
#    result is close to a random layout
fr_low_temp <- layout_with_fr(g, start.temp = 0.5)

# 6. Edge weights derived from graph-theoretic path distances.
#    weight = 1 / path_distance means short-path neighbours attract
#    more strongly and are pulled visually closer together.
path_dist    <- distances(g)             # all-pairs shortest paths
edge_list    <- as_edgelist(g, names = FALSE)
edge_weights <- 1 / path_dist[edge_list] # one weight per edge

fr_weighted <- layout_with_fr(g, weights = edge_weights)
```

### Plot the six FR variants

```{r}
#| label: fr-comparison
#| fig-cap: "Fruchterman-Reingold with varying `niter`, `start.temp`, and edge weights."
#| fig-height: 8

par(mfrow = c(2, 3), mar = c(1, 1, 3, 1))

plot(g, layout = fr_default,
     main         = "Default\n(niter = 500, default temp)",
     vertex.size  = node_size, vertex.color = node_col,
     vertex.label = NA, edge.color = edge_col, edge.width = edge_width)

plot(g, layout = fr_few_iter,
     main         = "Few iterations\n(niter = 30)",
     vertex.size  = node_size, vertex.color = node_col,
     vertex.label = NA, edge.color = edge_col, edge.width = edge_width)

plot(g, layout = fr_many_iter,
     main         = "Many iterations\n(niter = 5000)",
     vertex.size  = node_size, vertex.color = node_col,
     vertex.label = NA, edge.color = edge_col, edge.width = edge_width)

plot(g, layout = fr_high_temp,
     main         = "High temperature\n(start.temp = 2 × N)",
     vertex.size  = node_size, vertex.color = node_col,
     vertex.label = NA, edge.color = edge_col, edge.width = edge_width)

plot(g, layout = fr_low_temp,
     main         = "Low temperature\n(start.temp = 0.5)",
     vertex.size  = node_size, vertex.color = node_col,
     vertex.label = NA, edge.color = edge_col, edge.width = edge_width)

plot(g, layout = fr_weighted,
     main         = "Weighted edges\n(weight = 1 / path distance)",
     vertex.size  = node_size, vertex.color = node_col,
     vertex.label = NA, edge.color = edge_col, edge.width = edge_width)

par(mfrow = c(1, 1), mar = c(5, 4, 4, 2) + 0.1)
```

### What to notice

-   **Few iterations (niter = 30):** The layout looks incomplete — nodes have not had enough time to reach stable positions. Clusters are roughly visible but poorly resolved.
-   **Many iterations (niter = 5000):** The layout is more refined and reproducible, but for this network the visual difference from the default may be modest — the algorithm converges well before 5 000 steps.
-   **High temperature:** Nodes are spread aggressively across the canvas. Global cluster separation is exaggerated.
-   **Low temperature:** The result resembles a random layout because nodes never moved far from their initial random positions.
-   **Weighted edges:** Nodes that are close in graph-theoretic terms (short shortest path) are pulled tightly together visually, making dense local clusters compact and separation from the rest more obvious.

------------------------------------------------------------------------

## Summary

Layout algorithms are not neutral — each one privileges certain structural properties. Before choosing a layout:

1.  **Consider graph size.** FR and KK work well up to a few thousand nodes; DrL and LGL are designed for much larger graphs.
2.  **Consider what you want to show.** Geodesic distance → KK. Clusters → FR with enough iterations. Hierarchy → LGL. Pure enumeration → Circle.
3.  **Never use random as a final layout** — it is included here only as a visual baseline to show what "no information" looks like.
4.  **Tune FR parameters** when the default layout looks too crowded or too spread out. Start by adjusting `niter`, then `start.temp`.
