---
title: "R Basics"
subtitle: ""
author: "[Dr Clemens Jarnach](https://clemensjarnach.github.io)"
affiliation: "Oxford Martin School, University of Oxford"
date: today
format: 
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: cosmo
    self-contained: true
execute:
  warning: false
  message: false
---

This chapter introduces the R programming concepts you will need throughout the workshop. If you are already comfortable with R, treat it as a quick reference. If you are new to R, work through it carefully — the material here underpins everything that follows.

::: {.callout-tip}
## Running code

All code blocks in this workshop can be run interactively. Open the `.qmd` source file in RStudio, place your cursor inside a code chunk, and press **Ctrl+Enter** (Windows/Linux) or **Cmd+Enter** (Mac) to run a single line, or **Ctrl+Shift+Enter** to run the entire chunk.
:::

---

## Basic Operations and Assignment

R can be used as a calculator, but its real power lies in storing and manipulating values. The assignment operator `<-` binds a value to a name in the current environment.

```{r}
1 + 3           # arithmetic evaluation
2^8             # exponentiation
17 %% 5         # modulo (remainder)
17 %/% 5        # integer division

a <- 9          # assignment: bind 9 to the name 'a'
sqrt(a)         # apply a function to an object
b <- sqrt(a)

a == b          # logical comparison: are they equal?
a != b          # not equal
a > 5           # greater than

ls()            # list all objects in the current environment
```

::: {.callout-note}
R is case-sensitive: `A` and `a` are different objects. Object names can contain letters, digits, `.` and `_`, but must start with a letter or `.`.
:::

---

## Data Types

Every value in R has a type. The most common scalar types are:

```{r}
class(3.14)         # numeric (double)
class(42L)          # integer (note: I used the L suffix to tell R this is an integer)
class("hello")      # character
class(TRUE)         # logical
class(2 + 3i)       # complex

# Type coercion
as.numeric("3.14")
as.integer(42)
as.character(100)
as.logical(0)       # 0 is FALSE; anything non-zero is TRUE
is.na(NA)           # test for missing value
```

---

## Vectors

A vector is the fundamental data structure in R — an ordered collection of values of the **same type**. Almost all operations in R are vectorised, meaning they apply element-wise without explicit loops.

```{r}
x <- c(1, 3, 5)          # combine values into a vector
y <- c("one", "three", "five")

# Indexing (1-based)
x[1]                      # first element
x[c(1, 3)]                # first and third elements
x[-2]                     # all except the second

# Sequences
a <- 1:10
b <- seq(from = 0, to = 1, by = 0.25)
c_rep <- rep(c(1, 2), times = 3)
c_rep

# Vectorised operations — applied element-wise
x * 2
x + c(10, 20, 30)

# Logical operations on vectors
x > 2
any(x > 2)                # is any element > 2?
all(x > 2)                # are all elements > 2?
which(x > 2)              # indices where condition is TRUE

# Common summary functions
length(x)
sum(x)
mean(x)
sd(x)
min(x); max(x)
```

---

## Matrices

A matrix is a two-dimensional vector: all elements share the same type, and values are arranged in rows and columns. In network analysis, **adjacency matrices** are the canonical representation of graph structure — rows and columns correspond to nodes, and cell values indicate the presence (or weight) of an edge.

```{r}
m <- matrix(1:25, nrow = 5, ncol = 5)
m

# Element access
m[1, 2]                   # row 1, column 2
m[1, ]                    # entire first row
m[, 2]                    # entire second column
m[2:3, 3:5]               # submatrix

# Dimensions
nrow(m); ncol(m)
dim(m)

# Matrix operations — essential for network mathematics
t(m)                      # transpose
m %*% t(m)                # matrix multiplication (not element-wise!)
diag(m)                   # diagonal elements

# Build an adjacency matrix manually
adj <- matrix(0, nrow = 4, ncol = 4,
              dimnames = list(c("A","B","C","D"), c("A","B","C","D")))
adj["A", "B"] <- 1
adj["B", "C"] <- 1
adj["C", "D"] <- 1
adj["D", "A"] <- 1
adj                       # a 4-node directed cycle
```

---

## Lists

Lists are R's most flexible container: each element can hold any object of any type or length. Many R functions return lists (model output, igraph objects, etc.), so you need to know how to navigate them.

```{r}
person <- list(
  name    = "Alice",
  age     = 34,
  scores  = c(88, 92, 79),
  active  = TRUE
)

# Access by name
person$name
person[["age"]]

# Access by position
person[[3]]               # third element (the scores vector)
person[[3]][2]            # second score

# Inspect structure
str(person)
length(person)
names(person)
```

---

## Data Frames

A data frame is the standard rectangular data structure in R: a list of vectors of equal length, where each vector is a column. Think of it as a spreadsheet or database table. Node attribute tables and edge lists are both naturally represented as data frames.

```{r}
df <- data.frame(
  id     = 1:5,
  name   = c("Alice", "Bob", "Clara", "David", "Eva"),
  degree = c(3, 7, 2, 5, 4),
  active = c(TRUE, TRUE, FALSE, TRUE, FALSE),
  stringsAsFactors = FALSE
)

df

# Column access
df$name
df[["degree"]]

# Row and column indexing
df[1, ]                   # first row
df[, 3]                   # third column
df[df$active, ]           # rows where active == TRUE
df[df$degree > 3, c("name", "degree")]

# Useful inspection functions
nrow(df); ncol(df)
head(df, 3)
str(df)
summary(df)
```

---

## Control Flow

### Conditionals

```{r}
x <- 15

if (x > 10) {
  cat("x is greater than 10\n")
} else if (x == 10) {
  cat("x is exactly 10\n")
} else {
  cat("x is less than 10\n")
}

# ifelse() — vectorised conditional
scores <- c(55, 72, 48, 91, 60)
ifelse(scores >= 60, "pass", "fail")
```

### Loops

In R, explicit loops are often avoidable thanks to vectorisation, but they are useful for iterative tasks and are worth knowing.

```{r}
# for loop
for (i in 1:5) {
  cat("Iteration:", i, "\n")
}

# while loop
count <- 0
while (count < 3) {
  count <- count + 1
  cat("count is", count, "\n")
}

# break and next
for (i in 1:10) {
  if (i == 4) next       # skip this iteration
  if (i == 7) break      # exit the loop
  cat(i, "")
}
```

### Apply Functions

The `apply` family offers a concise way to apply a function over a vector, list, or matrix — often replacing explicit loops.

```{r}
m <- matrix(1:12, nrow = 3)

apply(m, 1, sum)          # row sums
apply(m, 2, mean)         # column means

nums <- list(a = 1:4, b = 5:10, c = 11:15)
sapply(nums, mean)        # returns a named vector
lapply(nums, length)      # returns a list
```

---

## Writing Functions

Functions are the building blocks of reusable code. When you find yourself repeating the same operations, write a function.

```{r}
# Basic function definition
greet <- function(name, greeting = "Hello") {
  paste(greeting, name)
}

greet("Alice")
greet("Bob", greeting = "Welcome")

# A more practical example: normalise a vector to [0, 1]
normalise <- function(x) {
  (x - min(x)) / (max(x) - min(x))
}

centralityScores <- c(3, 7, 2, 9, 4)
normalise(centralityScores)

# Functions can return multiple values via a list
describe <- function(x) {
  list(n = length(x), mean = mean(x), sd = sd(x), range = range(x))
}

describe(centralityScores)
```

---

## Data Wrangling with the Tidyverse

The [tidyverse](https://www.tidyverse.org) is a collection of R packages designed around a consistent philosophy of tidy data and readable code. The core package for data manipulation is **dplyr**, which provides a small set of verbs that cover the vast majority of data-wrangling tasks.

```{r}
#| message: false
#| warning: false

library(tidyverse)
```

### The Pipe Operator

The pipe `|>` (base R, since 4.1) or `%>%` (magrittr/tidyverse) passes the result of one expression as the first argument of the next. It allows you to write a sequence of transformations in the order they happen, which is much easier to read than nested function calls.

```{r}
# Without pipe — read inside out
round(mean(c(1, 2, 3, 4, 5)), digits = 2)

# With pipe — read left to right
c(1, 2, 3, 4, 5) |> mean() |> round(digits = 2)
```

### A Working Dataset

We'll use a small node-attribute table representing actors in a network.

```{r}
nodes <- tibble(
  id         = 1:8,
  name       = c("Alice","Bob","Clara","David","Eva","Frank","Grace","Hugo"),
  department = c("Research","Operations","Research","HR","Operations","Research","HR","Operations"),
  tenure     = c(5, 2, 8, 3, 6, 1, 4, 7),
  score      = c(72, 85, 91, 60, 78, 55, 88, 74)
)

nodes
```

### filter() — Keep Rows

```{r}
# Keep only Research department members
nodes |> filter(department == "Research")

# Multiple conditions
nodes |> filter(department == "Operations", tenure > 4)

# Using %in% for multiple values
nodes |> filter(department %in% c("HR", "Operations"))
```

### select() — Choose Columns

```{r}
nodes |> select(name, department, score)

# Drop a column with -
nodes |> select(-id)

# Rename while selecting
nodes |> select(name, dept = department, performance = score)
```

### mutate() — Create or Modify Columns

```{r}
nodes |>
  mutate(
    senior   = tenure >= 5,
    score_z  = (score - mean(score)) / sd(score),   # standardise
    label    = paste0(name, " (", department, ")")
  )
```

### arrange() — Sort Rows

```{r}
nodes |> arrange(desc(score))          # highest score first
nodes |> arrange(department, tenure)   # sort by two columns
```

### summarise() and group_by() — Aggregation

```{r}
# Overall summary
nodes |>
  summarise(
    n         = n(),
    mean_score = mean(score),
    sd_score   = sd(score),
    max_tenure = max(tenure)
  )

# Summary by group — critical for network attribute analysis
nodes |>
  group_by(department) |>
  summarise(
    n          = n(),
    mean_score = mean(score),
    mean_tenure = mean(tenure)
  ) |>
  arrange(desc(mean_score))
```

### Joining Tables

Joins combine two data frames on a shared key column — essential when merging node attributes with an edge list.

```{r}
# Edge list: pairs of node IDs representing connections
edges <- tibble(
  from = c(1, 1, 2, 3, 4, 5, 6),
  to   = c(2, 3, 4, 5, 6, 7, 8),
  weight = c(1.2, 0.8, 1.5, 0.5, 2.0, 1.1, 0.9)
)

# Attach sender attributes to edge list
edges |>
  left_join(nodes |> select(id, name, department),
            by = c("from" = "id")) |>
  rename(from_name = name, from_dept = department)
```

### Reshaping Data

```{r}
# Wide to long (pivot_longer) — useful for comparing multiple metrics
nodes |>
  select(name, tenure, score) |>
  pivot_longer(cols = c(tenure, score),
               names_to  = "metric",
               values_to = "value")

# Long to wide (pivot_wider)
nodes |>
  select(name, department) |>
  mutate(present = 1) |>
  pivot_wider(names_from = department, values_from = present, values_fill = 0)
```

---

## Reading and Writing Data

```{r}
#| eval: false

# CSV files
df <- read_csv("data/nodes.csv")            # tidyverse (recommended)
df <- read.csv("data/nodes.csv")            # base R

write_csv(df, "data/nodes_clean.csv")

# Excel files
library(readxl)
df <- read_excel("data/data.xlsx", sheet = 1)

# SPSS, Stata, SAS
library(haven)
df <- read_spss("data/survey.sav")
df <- read_dta("data/survey.dta")

# R's native format
saveRDS(df, "data/df.rds")
df <- readRDS("data/df.rds")
```

---

## Quick Reference

| Task | Function |
|---|---|
| Assign a value | `x <- value` |
| Create a vector | `c(1, 2, 3)` |
| Create a sequence | `1:10`, `seq(0, 1, 0.1)` |
| Create a matrix | `matrix(data, nrow, ncol)` |
| Create a data frame | `data.frame()` / `tibble()` |
| Check an object's type | `class()`, `typeof()` |
| Check dimensions | `dim()`, `nrow()`, `ncol()`, `length()` |
| Inspect structure | `str()`, `summary()`, `head()` |
| Filter rows | `filter()` |
| Select columns | `select()` |
| Create columns | `mutate()` |
| Sort rows | `arrange()` |
| Aggregate | `group_by()` + `summarise()` |
| Join tables | `left_join()`, `inner_join()` |
| Reshape wide → long | `pivot_longer()` |
| Reshape long → wide | `pivot_wider()` |


