---
title: "Social Network Analysis in R"
subtitle: "From Networks to Insights"
author: "Dr Clemens Jarnach"
affiliation: "Oxford Martin School, University of Oxford"
date: today
format: 
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: cosmo
    self-contained: true
execute:
  warning: false
  message: false
---

# Setup and Essentials

## Loading packages

```{r}
#| message: false
#| warning: false


library(igraph)
library(tidygraph)
library(ggraph)
library(tidyverse)
library(sna)
```

## Core R concepts

Before working with networks, we need a solid grasp of R fundamentals.

### Basic operations and assignment

```{r}
1 + 3           # evaluation
a <- 9          # assignment
sqrt(a)         # function application
b <- sqrt(a)
a == b          # logical comparison
ls()            # objects in environment
```

### Vectors

```{r}
x <- c(1, 3, 5)        # combine operator
x[2]                   # indexing
y <- c("one", "three", "five")

# Sequences
a <- 1:5
b <- seq(from = 1, to = 5, by = 1)
identical(a, b)

# Vectorised operations
x > 2
any(x > 2)
all(x > 2)
which(x > 2)
```

### Matrices

```{r}
m <- matrix(1:25, nrow = 5, ncol = 5)
m[1, 2]                 # element access
m[1, ]                  # row
m[, 2]                  # column
m[2:3, 3:5]             # submatrix

# Matrix operations
m %*% t(m)              # matrix multiplication
t(m)                    # transpose
dim(m)
```

### Data frames

```{r}
df <- data.frame(
  income = 1:5,
  sane = c(TRUE, TRUE, TRUE, TRUE, FALSE),
  name = LETTERS[1:5],
  stringsAsFactors = FALSE
)

df$sane                 # dollar notation
df[1, 2]
df[-1, ]                # negative indexing
```

---

# Importing and Creating Networks

## Built-in network data

```{r}
data(flo)               # Medici family network
flo
```

## Importing relational data

```{r}
# Read adjacency matrix
# relations <- read.csv("relationalData.csv", header = FALSE)
# relations <- as.matrix(relations)

# Read vertex attributes
# nodeInfo <- read.csv("vertexAttributes.csv", header = TRUE)

# Assign row/column names
# rownames(relations) <- nodeInfo$name
# colnames(relations) <- nodeInfo$name
```

## Creating network objects with igraph

The `igraph` package is now the standard for network analysis in R. It's faster, more flexible, and better maintained than older packages.

```{r}
# From adjacency matrix
g_flo <- graph_from_adjacency_matrix(flo, mode = "undirected", weighted = NULL)

# Quick inspection
g_flo
```

```{r}
# Create from edge list
edges <- tibble(
  from = c(1, 1, 2, 2, 3),
  to   = c(2, 3, 3, 4, 4)
)

g <- graph_from_data_frame(edges, directed = FALSE)
g
```

```{r}
# Empty network
g_empty <- make_empty_graph(n = 5, directed = FALSE)
```

---

# Network Description and Visualization

## Basic network properties

```{r}
vcount(g_flo)                  # number of vertices
ecount(g_flo)                  # number of edges
is_directed(g_flo)
is_weighted(g_flo)
```

## Static visualisation

### Base igraph plotting

```{r}
#| fig-width: 8
#| fig-height: 6
plot(g_flo, 
     vertex.size = 8,
     vertex.label.cex = 0.8,
     edge.width = 1.5)
```

### ggraph for publication-quality plots

```{r}
#| fig-width: 8
#| fig-height: 6
ggraph(g_flo, layout = 'fr') +
  geom_edge_link(edge_width = 0.5, edge_colour = 'grey70') +
  geom_node_point(size = 5, colour = 'steelblue') +
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  theme_graph()
```

---

# Working with Edges and Vertices

## Adding and removing edges

```{r}
g <- make_empty_graph(n = 5)

# Add edges by indexing
g <- g + edge(c(1, 2, 2, 3))
g


```

## Vertex and edge attributes

```{r}
# Add vertex attributes
V(g_flo)$degree <- igraph::degree(g_flo)
V(g_flo)
# Add edge attributes (on weighted example)
g_weighted <- g_flo
E(g_weighted)$weight <- runif(ecount(g_flo))
```

---

# Centrality Measures

## Degree centrality

```{r}
deg <- igraph::degree(g_flo)
deg
```

## Betweenness centrality

```{r}
bet <- igraph::betweenness(g_flo)
bet
```

## Closeness centrality

```{r}
clo <- igraph::closeness(g_flo)
clo
```

## Eigenvector centrality

```{r}
eig <- igraph::eigen_centrality(g_flo)$vector
eig
```

## Visualising centrality

```{r}
#| fig-width: 8
#| fig-height: 6
ggraph(g_flo, layout = 'fr') +
  geom_edge_link(edge_width = 0.5, edge_colour = 'grey70') +
  geom_node_point(aes(size = deg), colour = 'steelblue', alpha = 0.8) +
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  scale_size(range = c(2, 8)) +
  theme_graph() +
  labs(size = "Degree")
```

---

# Network-Level Indices

## Density and clustering

```{r}
edge_density(g_flo)
transitivity(g_flo)              # global clustering coefficient
```

## Reciprocity (directed networks)

```{r}
# Create a directed example
g_dir <- graph_from_data_frame(
  tibble(from = c(1, 2, 2, 3), to = c(2, 1, 3, 2)),
  directed = TRUE
)

reciprocity(g_dir)
```

## Components and connectivity

```{r}
igraph::components(g_flo)
is_connected(g_flo)
```
```{r}
# Geodesic distances
distances(g_flo)[1:3, 1:3]
diameter(g_flo)
```

---

# Subgroups and Cohesion

## k-cores

```{r}
kc <- coreness(g_flo)
kc
```

## Cliques

```{r}
cliques(g_flo, min = 3)
```

## Communities (modularity-based)

```{r}
comm <- cluster_louvain(g_flo)
membership(comm)
modularity(comm)
```

```{r}
#| fig-width: 8
#| fig-height: 6
ggraph(g_flo, layout = 'fr') +
  geom_edge_link(edge_width = 0.5, edge_colour = 'grey70') +
  geom_node_point(aes(colour = factor(membership(comm))), size = 5) +
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  theme_graph() +
  theme(legend.position = 'bottom')
```

---

# Structural Equivalence

```{r}
# Correlation of profiles
se_dist <- similarity(g_flo, method = "invlogweight")

# Hierarchical clustering
hc <- hclust(as.dist(1 - se_dist))
plot(hc)
```

---

# Random Graph Models

## Erdős–Rényi
Each possible edge is included independently with probability p. Produces a network with roughly uniform randomness and low clustering.
```{r}
g_er <- sample_gnp(n = 20, p = 0.15)

plot(g_er, main = "Erdős–Rényi (Gnp)", vertex.size = 20, vertex.label = NA, edge.color = "gray")
```

## Configuration model (fixed degree sequence)
Generates a random graph that preserves a specified degree sequence. Useful to test network properties while controlling for node degrees.
```{r}
g_conf <- igraph::sample_degseq(c(3,3,2,2,2))
plot(g_conf, main = "Configuration Model", vertex.size = 20, vertex.label = NA, edge.color = "steelblue")
```

## Small-world (Watts–Strogatz)
Starts with a regular lattice and randomly rewires edges with probability p. Produces networks with high clustering and short path lengths.

```{r}
g_ws <- sample_smallworld(dim = 1, size = 50, nei = 2, p = 0.1)
plot(g_ws, main = "Small-World (Watts–Strogatz)", vertex.size = 5, vertex.label = NA, edge.color = "orange")
```

## Scale-free (preferential attachment)
Nodes are added one by one, preferentially attaching to high-degree nodes. Produces networks with hubs and a power-law degree distribution.

```{r}
g_sf <- sample_pa(n = 50, power = 1, directed = FALSE)
plot(g_sf, main = "Scale-Free (Preferential Attachment)", vertex.size = 6, vertex.label = NA, edge.color = "purple")

```




---

# Further Resources

- **igraph documentation**: `?igraph`, `https://igraph.org/r/`
- **ggraph**: Extension for publication-quality plots
- **tidygraph**: Working with networks in the tidyverse
- **sna**: Classical functions (older but still useful)


