---
title: "Building Networks in R: A Step-by-Step Walkthrough"
subtitle: "Social Network Analysis Workshop"
author: "Dr Clemens Jarnach"
affiliation: "Oxford Martin School, University of Oxford"
date: "03-18-2026"
format: 
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: cosmo
    self-contained: true
execute:
  warning: false
  message: false
---

# Setup

Load the required packages:

```{r}
#| message: false
#| warning: false

library(igraph)
library(tidygraph)
library(ggraph)
library(tidyverse)
```

We'll build networks progressively, starting with three nodes and eventually adding structure, direction, and attributes.

------------------------------------------------------------------------

# Step 1: Three Isolated Nodes

Create a simple graph with three vertices and no edges yet.

```{r}
# Create vertices
nodes_1 <- tibble(
  id = 1:3,
  name = c("A", "B", "C")
)

# Create empty edge list
edges_1 <- tibble(
  from = integer(0),
  to = integer(0)
)

# Build the graph
g1 <- tbl_graph(nodes = nodes_1, edges = edges_1, directed = FALSE)

# Display
g1
```

Visualise the isolated nodes:

```{r}
#| fig-width: 8
#| fig-height: 6

ggraph(g1, layout = "fr") +
  geom_node_point(size = 8, colour = "steelblue") +
  geom_node_text(aes(label = name), repel = TRUE, size = 5) +
  theme_graph() +
  labs(title = "Step 1: Three Isolated Nodes")
```

------------------------------------------------------------------------

# Step 2: Adding Edges to the Nodes

Now we connect the three nodes with edges.

```{r}
# Define edges
edges_2 <- tibble(
  from = c(1, 2),
  to = c(2, 3)
)

# Build the graph
g2 <- tbl_graph(nodes = nodes_1, edges = edges_2, directed = FALSE)

g2
```

Visualise with edges:

```{r}
#| fig-width: 8
#| fig-height: 6

ggraph(g2, layout = "fr") +
  geom_edge_link(width = 1, colour = "grey40") +
  geom_node_point(size = 8, colour = "steelblue") +
  geom_node_text(aes(label = name), repel = TRUE, size = 5) +
  theme_graph() +
  labs(title = "Step 2: Adding Edges to the Nodes")
```

------------------------------------------------------------------------

# Step 3: Adding a Second Cluster

Introduce five additional nodes, densely connected among themselves but isolated from the first cluster.

```{r}
# Define all nodes (3 + 5)
nodes_3 <- tibble(
  id = 1:8,
  cluster = c(rep("A", 3), rep("B", 5))
)

# Define edges: dense connections within each cluster, none between
edges_3 <- tibble(
  from = c(
    # Cluster A: fully connected triangle
    1, 1, 2,
    # Cluster B: densely connected (not fully, for realism)
    4, 4, 4, 5, 5, 6
  ),
  to = c(
    # Cluster A
    2, 3, 3,
    # Cluster B
    5, 6, 7, 6, 8, 7
  )
)

# Build the graph
g3 <- tbl_graph(nodes = nodes_3, edges = edges_3, directed = FALSE)

g3
```

Visualise the two clusters:

```{r}
#| fig-width: 8
#| fig-height: 6

ggraph(g3, layout = "fr") +
  geom_edge_link(width = 1, colour = "grey40") +
  geom_node_point(aes(colour = cluster), size = 8) +
  geom_node_text(aes(label = id), repel = TRUE, size = 4) +
  scale_colour_manual(values = c("A" = "steelblue", "B" = "coral")) +
  theme_graph() +
  labs(title = "Step 3: Adding a Second Cluster")
```

Notice the structure: two dense groups with no connections between them.

------------------------------------------------------------------------

# Step 4: Bridging the Clusters

Add one edge connecting the two clusters. This single edge is a **bridge**.

```{r}
# Add one edge from cluster A to cluster B
edges_4 <- edges_3 %>%
  add_row(from = 3, to = 4)

# Build the graph
g4 <- tbl_graph(nodes = nodes_3, edges = edges_4, directed = FALSE)

g4
```

Visualise with the bridge:

```{r}
#| fig-width: 8
#| fig-height: 6

# Identify the bridge edge (the last one)
bridge_edge <- edges_4 %>% slice(n())

ggraph(g4, layout = "fr") +
  geom_edge_link(aes(colour = ifelse(from == 3 & to == 4, "bridge", "regular")), 
                 width = 1) +
  geom_node_point(aes(colour = cluster), size = 8) +
  geom_node_text(aes(label = id), repel = TRUE, size = 4) +
  scale_colour_manual(name = "Cluster", 
                      values = c("A" = "steelblue", "B" = "coral")) +
  scale_edge_colour_manual(name = "Edge Type",
                           values = c("bridge" = "darkred", "regular" = "grey40")) +
  theme_graph() +
  labs(title = "Step 4: Bridging the Clusters")
```

This single edge (node 3 to node 4) is critical for information flow between the two otherwise isolated groups.

------------------------------------------------------------------------

# Step 5: Introducing Direction

Add arrowheads to show direction. Convert to a directed graph.

```{r}
# Same edges, but now directed
g5 <- tbl_graph(nodes = nodes_3, edges = edges_4, directed = TRUE)

g5
```

Visualise as a directed graph:

```{r}
#| fig-width: 8
#| fig-height: 6

ggraph(g5, layout = "fr") +
  geom_edge_link(
    aes(colour = ifelse(from == 3 & to == 4, "bridge", "regular")),
    width = 1,
    arrow = arrow(length = unit(4, "mm"), type = "closed")
  ) +
  geom_node_point(aes(colour = cluster), size = 8) +
  geom_node_text(aes(label = id), repel = TRUE, size = 4) +
  scale_colour_manual(name = "Cluster",
                      values = c("A" = "steelblue", "B" = "coral")) +
  scale_edge_colour_manual(name = "Edge Type",
                           values = c("bridge" = "darkred", "regular" = "grey40")) +
  theme_graph() +
  labs(title = "Step 5: Introducing Direction")
```

In directed graphs, $i \to j$ is distinct from $j \to i$. The arrowheads show the direction of the relationship.

------------------------------------------------------------------------

# Step 6: Adding Node Attributes (Colour by Membership)

Colour nodes by cluster membership: one colour for cluster A, another for cluster B.

```{r}
# Nodes already have cluster attribute
nodes_3
```

Visualise with colour by cluster:

```{r}
#| fig-width: 8
#| fig-height: 6

ggraph(g5, layout = "fr") +
  geom_edge_link(
    aes(colour = ifelse(from == 3 & to == 4, "bridge", "regular")),
    width = 1,
    arrow = arrow(length = unit(4, "mm"), type = "closed")
  ) +
  geom_node_point(aes(colour = cluster), size = 10) +
  geom_node_text(aes(label = id), repel = TRUE, size = 4, colour = "white", fontface = "bold") +
  scale_colour_manual(name = "Cluster",
                      values = c("A" = "steelblue", "B" = "coral")) +
  scale_edge_colour_manual(name = "Edge Type",
                           values = c("bridge" = "darkred", "regular" = "grey40")) +
  theme_graph() +
  labs(title = "Step 6: Adding Node Attributes (Colour by Membership)")
```

Node attributes (colour, size, shape, etc.) encode information beyond the network structure itself. Here, colour reveals the cluster membership.

------------------------------------------------------------------------

# Summary

We've built a network progressively:

1.  **Three isolated nodes**: Starting point.
2.  **Add edges**: Connect nodes to form relationships.
3.  **Add a second cluster**: Introduce structure with two distinct groups.
4.  **Add a bridge**: Connect the clusters with a single critical edge.
5.  **Introduce direction**: Show asymmetric relationships with arrows.
6.  **Add node attributes**: Encode membership information via colour.

Each step reveals a different aspect of network structure and how to represent it in R.

------------------------------------------------------------------------

# Exercise

Try modifying the code:

-   Add more edges within each cluster. How does this change the visualisation?
-   Change the bridge from node 3→4 to node 2→5. Does the bridge still matter?
-   Add weights to edges. Hint: add a `weight` column to the edge list.
-   Create a third cluster and add multiple bridges connecting them.
